New orbits for irregular satellites based on thousands of new observations and predictions of stellar occultations up to 2020

### 2016-05-24 Irregular-occ/Irregular-occ.tex
.

\title[New orbits and predictions of stellar occultations by irregular satellites up to 2020]{New orbits for irregular satellites based on thousands of new observations and predictions of stellar occultations up to 2020}
\author[A. R. Gomes-J\'unior, M. Assafin, L. Beauvalet et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail: altair08@astro.ufrj.br},
M. Assafin$^{1,\dag}$,
L. Beauvalet$^{2,3}$,
J. Desmars$^{4}$,
R. Vieira-Martins$^{1,2,5,\dag}$,\newauthor
J. I. B. Camargo$^{2,5}$,
B. E. Morgado$^{1,2}$
F. Braga-Ribas$^{2,6}$
\\
$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
$^{2}$Observat\'orio Nacional/MCTI, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
$^{3}$Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France\\
$^{4}$Institut de m\'ecanique c\'eleste et de calcul des \'eph\'em\'erides - Observatoire de Paris, UMR 8028 du CNRS, 77 Av. Denfert-Rochereau,\\ 75014 Paris, France\\
$^{5}$Laborat\'orio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil\\
$^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
}
\begin{document}
\newcommand{\noccs}{5442 } % total number of occultations

@@

\cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas. For some satellites the number of positions obtained is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012}. They pointed out that the ephemeris of the irregular satellites have systematic errors that may reach 200 mas for some satellites.

We present in this paper new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) using only the positions obtained by \cite{GomesJunior2015}. We aim to eliminate the systematic errors in the JPL ephemeris (see Sec. \ref{Sec: integration}). %For an object at the distance of Jupiter, this represents an error langer than 700 km in the shadow path of an occultation.
For Phoebe, we updated the ephemeris of \cite{Desmars2013} using the observations of \cite{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Center and observations from Flagstaff.

@@

\cite{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.

Here we compute new orbits based on the observations published in \cite{GomesJunior2015}. First, because the reduction was made with a consistent and precise stellar catalogue and with a robust astrometry \citep[PRAIA,][]{Assafin2011}. Second, besides recent observations, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets. For these reasons, only this set of positions was used for the satellites of Jupiter.

@@

Fig. \ref{Fig: JPL-STE} displays the offsets of the positions published by \cite{GomesJunior2015} for the satellite Carme in declination relative to the STE ephemeris, to \cite{Jacobson2012} JUP300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.We see that the systematic JPL ephemeris offsets pointed out by \cite{GomesJunior2015} are reduced with our ephemeris, as expected.

\begin{figure}
\includegraphics[width=8.8cm]{figures/Carme_ephemeris.eps}
\caption{Offsets in declination of the positions published by \protect\cite{GomesJunior2015} for Carme. The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the JUP300 JPL ephemeris and the green dot relative to \protect\cite{Emelyanov2008}. As expected, the ephemeris systematic errors pointed out by \protect\cite{GomesJunior2015} are reduced with the STE ephemeris. \label{Fig: JPL-STE}}
\end{figure}

Table \ref{Tab:STE-error}

\begin{table}
\caption{\label{Tab:STE-error} Mean offsets of the \protect\cite{GomesJunior2015} positions relative to the STE, \protect\cite{Jacobson2012} and \protect\cite{Emelyanov2008} ephemeris .}
\begin{centering}
\begin{tabular}{lcccccc}
\hline  \hline
%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
 & \multicolumn{2}{c}{STE} & \multicolumn{2}{c}{JPL} & \multicolumn{2}{c}{Eme2008}\tabularnewline
Satellite  & RA & DEC & RA & DEC & RA & DEC \tabularnewline
\hline
Ananke & 12 & 16 & 49 & 359 & 187 & 623 \tabularnewline
Carme & 20 & 14 & 30 & 369 & 220 & 653 \tabularnewline
Elara & 14 & 16 & 33 & 305 & 193 & 561 \tabularnewline
Himalia & 15 & 12 & 54 & 257 & 230 & 568 \tabularnewline
Leda & 8 & 24 & 38 & 362 & 208 & 640 \tabularnewline
Lysithea & 16 & 11 & 35 & 330 & 212 & 604 \tabularnewline
Pasiphae & 20 & 19 & 44 & 362 & 206 & 651 \tabularnewline
Sinope & 15 & 21 & 34 & 356 & 256 & 682 \tabularnewline
\hline
\end{tabular}
\par \end{centering}
\end{table}

@@

For the specific case of Phoebe, the ninth satellite of Saturn, we have updated the ephemeris published in \cite{Desmars2013}. The new ephemeris (PH15) used the same dynamical model, including the perturbations of the Sun and the eight planets, the eight major satellites of Saturn and the $J_2$ parameter. The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \cite{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center\footnote{\url{http://lnfm1.sai.msu.ru/neb/nss/bsapoouf.htm}}), and observations from Flagstaff \citep{NOFS} between 2012 and 2014. It represents a total number of 5886 observations from 1898 to 2014. In contrast, in \cite{Desmars2013} was used 3367 observations from 1898 to 2012. It represents an increase of almost 75\% in the number of observations.

@@

@@

The first preliminary catalogue version of the ESA astrometry satellite GAIA \citep{deBruijne2012} is expected to be released up to the end of 2016 (The catalogue with five-parameter astrometric solutions is up to the end of 2017). The precise star positions to be derived by GAIA will provide better predictions with the main source of error being the ephemeris. Astrometric reduction of observations published in \cite{GomesJunior2015} will be revised with the GAIA catalogue and the predictions will be improved. Since we cannot foresee exactly when the GAIA catalogue will be released and when new and re-reduced satellite positions in the GAIA frame will become available, we decided to publish predictions up to the end of 2020. In the GAIA era, the occultations predicted will be updated.

@@

For the irregular satellites of Jupiter (Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope), we produced ephemeris using only the observations of \cite{GomesJunior2015}. The systematic errors found in the JPL ephemeris \citep{Jacobson2012} and \cite{Emelyanov2008} were corrected. This new ephemeris was denominated Special Tailored Ephemeris.

We also updated the ephemeris of Phoebe \citep{Desmars2013} using the observations of \cite{GomesJunior2015}, \cite{Peng2015} observations from  MPC and from Flagstaff. A total of 5886 observations between 1989 and 2014 were used in the process. This represents and increase of about 75\% in the number used to generate the orbits of Phoebe by \cite{Desmars2013}.

\textcolor{red}{Colocar mais alguma coisa sobre as efemérides}

@@

\cite{GomesJunior2015} also observed Sycorax (satellite of Uranus) and Nereid (satellite of Neptune). Sycorax had a few observations on 9 nights in two years which did not cover one orbital period. For Nereid, the observations covered many orbital periods, but due to Nereid's large orbital eccentricity there are no observations near the pericenter.

>>>>

\title[New orbits and predictions of stellar occultations by irregular satellites up to 2020]{New orbits for irregular satellites based on thousands of new observations and predictions of stellar occultations up to 2020}
\author[A. R. Gomes-J\'unior, M. Assafin, L. Beauvalet et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail: altair08@astro.ufrj.br},
M. Assafin$^{1,\dag}$,
L. Beauvalet$^{2,3}$,
J. Desmars$^{4}$,
R. Vieira-Martins$^{1,2,5,\dag}$,\newauthor
J. I. B. Camargo$^{2,5}$,
B. E. Morgado$^{1,2}$
F. Braga-Ribas$^{2,6}$
\\
$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
$^{2}$Observat\'orio Nacional/MCTI, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
$^{3}$Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France\\
$^{4}$Institut de m\'ecanique c\'eleste et de calcul des \'eph\'em\'erides - Observatoire de Paris, UMR 8028 du CNRS, 77 Av. Denfert-Rochereau,\\ 75014 Paris, France\\
$^{5}$Laborat\'orio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil\\
$^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
}
\begin{document}
\newcommand{\noccs}{5442 } % total number of occultations
\defcitealias{GomesJunior2015}{GJ15}

@@

\cite{GomesJunior2015} (which shall be referred as \citetalias{GomesJunior2015} hereafter) obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas. For some satellites the number of positions obtained is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012}. They pointed out that the ephemeris of the irregular satellites have systematic errors that may reach 200 mas for some satellites.

We present in this paper new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) using only the positions obtained by \citetalias{GomesJunior2015}. We aim to eliminate the systematic errors in the JPL ephemeris (see Sec. \ref{Sec: integration}). %For an object at the distance of Jupiter, this represents an error langer than 700 km in the shadow path of an occultation.
For Phoebe, we updated the ephemeris of \cite{Desmars2013} using the observations of \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Center and observations from Flagstaff.

@@

\citetalias{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.

Here we compute new orbits based on the observations published in \citetalias{GomesJunior2015}. First, because the reduction was made with a consistent and precise stellar catalogue and with a robust astrometry \citep[PRAIA,][]{Assafin2011}. Second, besides recent observations, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets. For these reasons, only this set of positions was used for the satellites of Jupiter.

@@

Fig. \ref{Fig: JPL-STE} displays the offsets of the positions published by \citetalias{GomesJunior2015} for the satellite Carme in declination relative to the STE ephemeris, to \cite{Jacobson2012} JUP300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.We see that the systematic JPL ephemeris offsets pointed out by \citetalias{GomesJunior2015} are reduced with our ephemeris, as expected.

\begin{figure}
\includegraphics[width=8.8cm]{figures/Carme_ephemeris.eps}
\caption{Offsets in declination of the positions published by \protect\citetalias{GomesJunior2015} for Carme. The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the JUP300 JPL ephemeris and the green dot relative to \protect\cite{Emelyanov2008}. As expected, the ephemeris systematic errors pointed out by \protect\citetalias{GomesJunior2015} are reduced with the STE ephemeris. \label{Fig: JPL-STE}}
\end{figure}

In Table \ref{Tab:STE-error} it is presented the mean offsets of the \citetalias{GomesJunior2015} positions relative to the same 3 ephemeris as above. We can see that the mean offsets are highly reduced with the STE ephemeris.

\begin{table}
\caption{\label{Tab:STE-error} Mean offsets of the \protect\citetalias{GomesJunior2015} positions relative to the STE, \protect\cite{Jacobson2012} and \protect\cite{Emelyanov2008} ephemeris .}
\begin{centering}
\begin{tabular}{lcccccc}
\hline  \hline
%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
 & \multicolumn{2}{c}{STE} & \multicolumn{2}{c}{JPL} & \multicolumn{2}{c}{Eme2008}\tabularnewline
Satellite  & RA & DEC & RA & DEC & RA & DEC \tabularnewline
\hline
Ananke & $-10$ & $3$ & $60$ & $-108$ & $101$ & $-107$ \tabularnewline
Carme & $14$ & $-1$ & $-3$ & $-80$ & $-6$ & $-45$ \tabularnewline
Elara & $3$ & $-12$ & $20$ & $-50$ & $23$ & $-83$ \tabularnewline
Himalia & $-15$ & $-7$  & $-19$  & $-11$ & $-18$  & $-13$ \tabularnewline
Leda & $-9$ & $-8$ & $60$ & $-13$ & $166$ & $92$ \tabularnewline
Lysithea & $15$ & $-21$ & $40$ & $-43$ & $117$ & $-76$ \tabularnewline
Pasiphae & $4$ & $-16$ & $-17$ & $-82$ & $-10$ & $-54$ \tabularnewline
Sinope & $9$ & $-4$ & $10$ & $-35$ & $11$ & $-52$ \tabularnewline
\hline
\end{tabular}
\par \end{centering}
\end{table}

@@

For the specific case of Phoebe, the ninth satellite of Saturn, we have updated the ephemeris published in \cite{Desmars2013}. The new ephemeris (PH15) used the same dynamical model, including the perturbations of the Sun and the eight planets, the eight major satellites of Saturn and the $J_2$ parameter. The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center\footnote{\url{http://lnfm1.sai.msu.ru/neb/nss/bsapoouf.htm}}), and observations from Flagstaff \citep{NOFS} between 2012 and 2014. It represents a total number of 5886 observations from 1898 to 2014. In contrast, in \cite{Desmars2013} was used 3367 observations from 1898 to 2012. It represents an increase of almost 75\% in the number of observations.

@@

We computed the precision of the PH15 ephemeris considering three sets of observations: all the positions available; only the positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}. In Fig. \ref{Fig:err-Phoebe} is shown the comparison between them for the time spam 2016-2021. It is possible to see that even considering only the positions of \citetalias{GomesJunior2015}, the estimated error of the ephemeris is smaller than 12 mas. The computed precision does not take into account the precision in the position of Saturn.

\begin{figure}
\begin{centering}
\includegraphics[width=8.8cm]{figures/PH15_err_angsep.eps}  
\caption{Comparison of the precision of the PH15 ephemeris where three different sets of positions were used to compute the ephemeris: all observations; only positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}.}
\label{Fig:err-Phoebe}
\end{centering}
\end{figure}

@@

The first preliminary catalogue version of the ESA astrometry satellite GAIA \citep{deBruijne2012} is expected to be released up to the end of 2016 (The catalogue with five-parameter astrometric solutions is up to the end of 2017). The precise star positions to be derived by GAIA will provide better predictions with the main source of error being the ephemeris. Astrometric reduction of observations published in \citetalias{GomesJunior2015} will be revised with the GAIA catalogue and the predictions will be improved. Since we cannot foresee exactly when the GAIA catalogue will be released and when new and re-reduced satellite positions in the GAIA frame will become available, we decided to publish predictions up to the end of 2020. In the GAIA era, the occultations predicted will be updated.

@@

For the irregular satellites of Jupiter (Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope), we produced ephemeris using only the observations of \citetalias{GomesJunior2015}. The systematic errors found in the JPL ephemeris \citep{Jacobson2012} and \cite{Emelyanov2008} were corrected. This new ephemeris was denominated Special Tailored Ephemeris.

We also updated the ephemeris of Phoebe \citep{Desmars2013} using the observations of \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from  MPC and from Flagstaff. A total of 5886 observations between 1989 and 2014 were used in the process. This represents and increase of about 75\% in the number used to generate the orbits of Phoebe by \cite{Desmars2013}.

As it was shown for Phoebe, when we use only the positions of \citetalias{GomesJunior2015}, the ephemeris presents a precision in the order of those where all the positions were used. Moreover, the case of Phoebe is particular because we have many observations in a large time spam (1898 to 2014) including observations from Cassini. For the Jovian satellites with less observations, the precision we have by using only \citetalias{GomesJunior2015} observations would be quite equivalent or even better than the precision of other ephemeris for the short time explored in this work.

@@

\citetalias{GomesJunior2015} also observed Sycorax (satellite of Uranus) and Nereid (satellite of Neptune). Sycorax had a few observations on 9 nights in two years which did not cover one orbital period. For Nereid, the observations covered many orbital periods, but due to Nereid's large orbital eccentricity there are no observations near the pericenter.
